---
title: "base_analysis"
author: "Margot Lautens"
updated: '2023-11-13'
---

# Summary

The goal of this program is to process LC-MS data from output chromXtractor by first removing low quality peaks then normalizing and finding which peaks and pathways are significant for your interaction of interest. Plots and heatmaps can also be generated.

# Setup

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = getwd()) #Sets the current directory as your working directory.

# Required libraries
library(tidyverse) # Required for syntax used in writing the function
library(readr) # Required for easy loading

experiment_name <- "example"
```

## Load Required Datasets
```{r}
sample_sheet <- read.csv("data/base_sample_sheet.csv")%>%
  filter(KCN!="1KCN")
View(sample_sheet)

compounds <- read.csv("data/base_compounds.csv")%>%
    mutate(KEGG_ID=gsub("C","",KEGG_ID))
View(compounds)

objectDataFrame <- read.csv("data/base_dataset2.csv",
                            sep=";")%>%
  mutate(oldSampleName=sampleName)%>%
  mutate(sampleName=gsub("20181010_","",sampleName))%>%
  mutate(sampleName=gsub(".d","",sampleName))
View(objectDataFrame)
```

## Download or Load a Pathway File
```{r}
library(readxl)
cel_model <- read_excel("data/iCEL1314.xlsx")%>%
  rename(HRF=`Human readable formula`)%>%
  rename(MRF=`Machine readable formula`) # If using an existing file.

#source('scripts/keggExtractor.R') # To generate the list of KEGG pathways

#cel_model <- keggExtractor("cel") # If downloading from KEGG.

View(cel_model)
```

# Clean and QC Data
```{r, warning=FALSE}
source('scripts/dataCleanup.R')  # For data cleanup functions.

sampleData <- dataCleanup(objectDataFrame,
                          sampleTable=sample_sheet, 
                          compounds=compounds,
                          # Name the way you're measuring your peak
                          peak=integratedIntensity,
                          # Select the variables of interest (ie the drugs or mutants)
                          factors=c("Mutant","KCN"),
                          # 'cut_height' governs the minimum intensity of a peak.
                          cut_height=1250,
                          # 'percentile' governs the cutoff for how what percentage of a peak a                            blank ('percentile') or a filter ('percentile2') can be relative to a                           peak in a sample. 1 is the most permissive and '0' the least.
                          percentile=0.9,percentile2=0.9,
                          # If you have an internal reference, this will normalize each peak to                             that reference
                          refNorm=TRUE,
                          KEGG=TRUE)

print(filteredMets)
View(sampleData)

# 'filteredMets' are the metabolites which were removed for failing to meet the quality standard.
```

## Comments:

## This function keeps features which have peaks beyond the set cutoff (percentile) in regards to the blank or filter for that peak.

## The peaks can also be normalized to the internal reference peak.

## This function will save a 'runbanks' and a 'repblanks' histogram to your working directory.

## These show a histogram of the peaks where x is the ratio between the sample and the blank/filter. Check that you are happy with your select cutoff (the line) before continuing.

## The peaks which were below the quality threshold as stored in 'filteredMets' and will be printed for you. Be sure to check if you're missing any important favourites and adjust your 'percentile's as needed.

```{r}
## !!! Important !!
## Must modify for your normalizations of interest.
source('scripts/base_Normalization_Scripts.R') 
View(fn_model)

pVal=0.05 

# This will normalize your data in each of the ways you list in your normalization scripts so that you can compare normalization methods.
normData <- list_model%>%
  map_df(fn_model,sampleData%>%
           nest(data=everything()),
         .id="normMethod")%>%
  unnest(normData)%>%
  mutate(saveName=paste0(saveName,pVal,"_",normMethod,"_",featureName,"_"))%>%
              mutate(experiment=experiment_name)

write.table(normData%>%
              select(-factors,-data), paste0(saveName,"norm.csv"),
            quote=FALSE, sep=";", row.names=FALSE)
#View(normData)
```

## Comments:

## You must modify the 'base_Normalization_Scripts' yourself for every project. This is where you choose which types of normalization you'd like to run. You need to write at least one but you can write as many as you'd like.

## Must include:

## - Within each function:

## - A 'factors' column

## - A 'form' column

## - 'list_model'

## - 'fn_model'

## A full guide is within the script.

## 'normData' will be your normalized data which will used throughout the following steps.

```{r, warning=FALSE}
source('scripts/analysisTools2.R') # To find significant peaks and plot them

sign_mets <- signANOVA(normData,
                       sampleData=sampleData,
                       met_model = cel_model,
                       saveName=saveName,
                       pVal=pVal,
                       target="1KCN:1Rot-1KCN:0Rot" ,
                       #This is to determine what the target comparison you're interested in is.
                       show_all=TRUE
                       )

```

## Comments:

## This will generate both a 'volcanoPlot.png' and 'sign_mets'.

## The volcano plot is a spread of the difference for each peak between different factors and terms on the x axis and the p value for each of those differences for each peak on the y. A subplot is made for each type of normalization and for each term ie in a two factor linear model of foldChange\~factor1\*factor2, the terms will be 'factor1', 'factor2' and 'factor1:factor2'. The points in blue are those which go beyond your 'pVal' cutoff for both the x and the y axis. The number in the right hand corner is the number of peaks which meet that cutoff. This is one way of comparing the different normalization options that you have selected.

## When you're happy with your normalizations, continue.

## 'sign_mets' is a dataframe of all the metabolites which passed your cutoff for each individual term:normalization interaction. The KEGG pathway(s) which those metabolites belong to are also listed.

```{r}
source("scripts/Heatmap_code.R") # To make heatmaps.

heatmap_table <- makeHeatmapTable(met_data = normData,
                                  # A table of features and nomalizations of interest as in 
                                  # 'sign_mets', optional.
                                  #met_table = sign_mets,
                                  # A vector of metabolites of interest, optional.
                                  #met_list = c("L_Tryptophan","L_Malic_Acid"),
                                  # A vector of normalizations of interest, optional.
                                  #norm_list = c("mutantnorm"),
                                  value_y = "foldChange",
                                  # Do you want each rep or a mean of the reps.
                                  data_mean = TRUE,
                                  # Whether the data will be sorted by level or by clustering.
                                  data_sort = TRUE,
                                  filetype=".pdf",
                                  same_legend = T,
                                  sym_breaks = T
                                  )
```

## Comments:

## This will generate a table of heatmaps for each normalization as well as saving them.

## If desired, you can select a table (like 'sign_mets') to filter your data. A vector of metabolites or of normalizations can also be used to filter the data used to generate the heatmap.

## 'y' must be a string.

```{r}
source('scripts/analysisTools2.R') # To find significant peaks and plot them

plots1 <- plotPeaks(normData,
# The y axis variable.
                    yPlot="foldChange",
                    xPlot="KCN",
# The variable which will be used to color the boxplots. Must be one of the factors. If one of the factors you are analyzing is a mutant/strain, you must use that as the fill.
                    fillPlot="Rot",
# A vector of the labels for your fill variable.
                    fillVector=c("0.8% DMSO","12.5uM Rot"),
# The S4 object output of chromeXtractor, necessary is you would like to plot the peaks.
                    integratedset=integratedset,
                    samples=samples,
                    sampleData=sampleData,
                    met_model=cel_model,
# Whether to plot the boxplots for each feature.
                    plotPlot=TRUE,
# Whether to plot the pvalue brackets on the boxplots.
                    plotP=TRUE,
# Whether to plot the mass spec peak for each feature.
                    plotPeak=F)

#save.image(paste0(saveName,".RData")) # Use if you'd like to save the plotData (useful for version control but will take a while to save and to open).
```

## Comments:

## This will generate a table with boxplots and/or peaks for each feature and normalization method. To see a specific graph just select the appropriate row and print 'savedPlot' or 'savedPeak'. You can also 'pull()' peaks/plots of interest.

```{r}

View(sign_mets%>%
    ungroup()%>%
    count(normMethod,Pathway)%>%
    arrange(desc(n)))

```

## Comments:

## This is just a quick piece of code to see which pathways are enriched in the sign_mets which is helpful for the next step. There is no correction so be sure to use another program if you're interested in proper pathway enrichment.

```{r}

source('Scripts/analysisTools3.R') # To save plots.

savePeaks(plots1,
          final.cmpd.df,
# A vector of pathways of interest, optional.
          #pathway_list = c("Tryptophan metabolism"),
# A vector of metabolites of interest, optional. 
          #met_list = rot_sign_mets%>%
    #select(-term,-normMethod),
# A vector of normalizations of interest, optional.
          norm_list = c("mutantnorm"),
# Whether to save the peak for each feature.
          savePeak = F,
# Whether to save the plot for each feature.
          savePlot = T,
# Specific filepath other than the working directory, must end with a '/', optional.
          #path="plots/",
# Filetype (ie. '.svg', '.png', '.jpg'), optionally add any additional text for the filename.
          fileType = ".svg")
```

## Comments:

## Saves peaks and boxplots.
